---
title: "scRNA-Seq Lab"
author: "Matt Mueller"
date: "2024-04-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE}
options(warn = -1)
suppressMessages(library(Seurat))
suppressMessages(library(tidyverse))
suppressMessages(library(Matrix))
suppressMessages(library(RCurl))
suppressMessages(library(scales))
suppressMessages(library(cowplot))
suppressMessages(library(metap))
suppressMessages(library(AnnotationHub))
suppressMessages(library(ensembldb))
suppressMessages(library(multtest))
suppressMessages(library(glmGamPoi))
```

```{r}
library(Seurat)
library(tidyverse)
library(Matrix)
library(RCurl)
library(scales)
library(cowplot)
library(metap)
library(AnnotationHub)
library(ensembldb)
library(multtest)
library(glmGamPoi)


#### Data Quality Control and Clustering/PCA ####

# How to read in 10X data for a single sample (output is a sparse matrix)
ctrl_counts <- Read10X(data.dir = "C:/Users/13145/OneDrive/Desktop/BCB 5250/single_cell_rnaseq/single_cell_rnaseq/data/ctrl_raw_feature_bc_matrix")

# Turn count matrix into a Seurat object (output is a Seurat object)
ctrl <- CreateSeuratObject(counts = ctrl_counts,
                           min.features = 100)

# Explore the metadata
head(ctrl@meta.data)

# Create a Seurat object for each sample
for (file in c("ctrl_raw_feature_bc_matrix", "stim_raw_feature_bc_matrix")){
  seurat_data <- Read10X(data.dir = paste0("C:/Users/13145/OneDrive/Desktop/BCB 5250/single_cell_rnaseq/single_cell_rnaseq/data/", file))
  seurat_obj <- CreateSeuratObject(counts = seurat_data, 
                                   min.features = 100, 
                                   project = file)
  assign(file, seurat_obj)
}

# Check the metadata in the new Seurat objects
head(ctrl_raw_feature_bc_matrix@meta.data)
head(stim_raw_feature_bc_matrix@meta.data)
```
##  Question: What is min.features? Describe it and min.features = 100. 
#     Min.features sets the minimum number of a specified feature needed to include the data point in analysis. In this case, min.features = 100 means that a minimum of 100 detected genes is needed to include a cell in analysis.

##  Question: Report nCount_RNA and nFeature_RNA for AAACATACATTTCC-1 (2nd item) from the ctrl_raw_feature_bc_matrix

# nCount_RNA is 3125 and nFeature_RNA is 896.

```{r}
# Create a merged Seurat object
merged_seurat <- merge(x = ctrl_raw_feature_bc_matrix, 
                       y = stim_raw_feature_bc_matrix, 
                       add.cell.id = c("ctrl", "stim"))

# Check that the merged object has the appropriate sample-specific prefixes
head(merged_seurat@meta.data)
tail(merged_seurat@meta.data)

# Explore merged metadata
View(merged_seurat@meta.data)

# Add number of genes per UMI for each cell to metadata
merged_seurat$log10GenesPerUMI <- log10(merged_seurat$nFeature_RNA) / log10(merged_seurat$nCount_RNA)

# Compute percent mito ratio
merged_seurat$mitoRatio <- PercentageFeatureSet(object = merged_seurat, pattern = "^MT-")
merged_seurat$mitoRatio <- merged_seurat@meta.data$mitoRatio / 100

# Create metadata dataframe
metadata <- merged_seurat@meta.data

# Add cell IDs to metadata
metadata$cells <- rownames(metadata)

# Create sample column
metadata$sample <- NA
metadata$sample[which(str_detect(metadata$cells, "^ctrl_"))] <- "ctrl"
metadata$sample[which(str_detect(metadata$cells, "^stim_"))] <- "stim"

# Rename columns
metadata <- metadata %>%
  dplyr::rename(seq_folder = orig.ident,
                nUMI = nCount_RNA,
                nGene = nFeature_RNA)

# Add metadata back to Seurat object
merged_seurat@meta.data <- metadata

# Create .RData object to load at any time
save(merged_seurat, file="C:/Users/13145/OneDrive/Desktop/BCB 5250/single_cell_rnaseq/single_cell_rnaseq/data/merged_filtered_seurat.RData")

# Visualize the number of cell counts per sample
metadata %>% 
  ggplot(aes(x=sample, fill=sample)) + 
  geom_bar() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  theme(plot.title = element_text(hjust=0.5, face="bold")) +
  ggtitle("NCells")

# Visualize the number UMIs/transcripts per cell
metadata %>% 
  ggplot(aes(color=sample, x=nUMI, fill= sample)) + 
  geom_density(alpha = 0.2) + 
  scale_x_log10() + 
  theme_classic() +
  ylab("Cell density") +
  geom_vline(xintercept = 500)

# Visualize the distribution of genes detected per cell via histogram
metadata %>% 
  ggplot(aes(color=sample, x=nGene, fill= sample)) + 
  geom_density(alpha = 0.2) + 
  theme_classic() +
  scale_x_log10() + 
  geom_vline(xintercept = 300)

# Visualize the overall complexity of the gene expression by visualizing the genes detected per UMI (novelty score)
metadata %>%
  ggplot(aes(x=log10GenesPerUMI, color = sample, fill=sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  geom_vline(xintercept = 0.8)

# Visualize the distribution of mitochondrial gene expression detected per cell
metadata %>% 
  ggplot(aes(color=sample, x=mitoRatio, fill=sample)) + 
  geom_density(alpha = 0.2) + 
  scale_x_log10() + 
  theme_classic() +
  geom_vline(xintercept = 0.2)

# Visualize the correlation between genes detected and number of UMIs and determine whether strong presence of cells with low numbers of genes/UMIs
metadata %>% 
  ggplot(aes(x=nUMI, y=nGene, color=mitoRatio)) + 
  geom_point() + 
  scale_colour_gradient(low = "gray90", high = "black") +
  stat_smooth(method=lm) +
  scale_x_log10() + 
  scale_y_log10() + 
  theme_classic() +
  geom_vline(xintercept = 500) +
  geom_hline(yintercept = 250) +
  facet_wrap(~sample)

# Filter out low quality cells using selected thresholds - these will change with experiment
filtered_seurat <- subset(x = merged_seurat, 
                          subset= (nUMI >= 500) & 
                            (nGene >= 250) & 
                            (log10GenesPerUMI > 0.80) & 
                            (mitoRatio < 0.20))

# Extract counts
counts <- GetAssayData(object = filtered_seurat, slot = "counts.stim_raw_feature_bc_matrix")

# Output a logical matrix specifying for each gene on whether or not there are more than zero counts per cell
nonzero <- counts > 0

# Sums all TRUE values and returns TRUE if more than 10 TRUE values per gene
keep_genes <- Matrix::rowSums(nonzero) >= 10

# Only keeping those genes expressed in more than 10 cells
filtered_counts <- counts[keep_genes, ]

# Reassign to filtered Seurat object
filtered_seurat <- CreateSeuratObject(filtered_counts, meta.data = filtered_seurat@meta.data)

# Save filtered subset to new metadata
metadata_clean <- filtered_seurat@meta.data

# Normalize the counts
seurat_phase <- NormalizeData(filtered_seurat)

# Load cell cycle markers
load("C:/Users/13145/OneDrive/Desktop/BCB 5250/single_cell_rnaseq/single_cell_rnaseq/data/cycle.rda")

# Score cells for cell cycle
seurat_phase <- CellCycleScoring(seurat_phase, 
                                 g2m.features = g2m_genes, 
                                 s.features = s_genes)

# View cell cycle scores and phases assigned to cells                                 
View(seurat_phase@meta.data)  

# Identify the most variable genes
seurat_phase <- FindVariableFeatures(seurat_phase, 
                                     selection.method = "vst",
                                     nfeatures = 2000, 
                                     verbose = FALSE)

# Scale the counts
seurat_phase <- ScaleData(seurat_phase)

# Identify the 15 most highly variable genes
ranked_variable_genes <- VariableFeatures(seurat_phase)
top_genes <- ranked_variable_genes[1:15]

# Plot the average expression and variance of these genes
# With labels to indicate which genes are in the top 15
p <- VariableFeaturePlot(seurat_phase)
LabelPoints(plot = p, points = top_genes, repel = TRUE)

# Perform PCA
seurat_phase <- RunPCA(seurat_phase)

# Plot the PCA colored by cell cycle phase
DimPlot(seurat_phase,
        reduction = "pca",
        group.by= "Phase",
        split.by = "Phase")

# Check quartile values
summary(seurat_phase@meta.data$mitoRatio)

# Turn mitoRatio into categorical factor vector based on quartile values
seurat_phase@meta.data$mitoFr <- cut(seurat_phase@meta.data$mitoRatio, 
                                     breaks=c(-Inf, 0.0144, 0.0199, 0.0267, Inf), 
                                     labels=c("Low","Medium","Medium high", "High"))
# PCA plot for mitoFR
DimPlot(seurat_phase,
        reduction = "pca",
        group.by= "mitoFr",
        split.by = "mitoFr")


# Split seurat object by condition to perform cell cycle scoring and SCT on all samples
split_seurat <- SplitObject(seurat_phase, split.by = "sample")

# adjust limit for allowable object sizes
options(future.globals.maxSize = 4000 * 1024^2)
```

```{r, echo=FALSE}
# perform the sctransform on all samples
for (i in 1:length(split_seurat)) {
  split_seurat[[i]] <- SCTransform(split_seurat[[i]], vars.to.regress = c("mitoRatio"), vst.flavor = "v2")
}
```

```{r}
# Check which assays are stored in objects
split_seurat$stim@assays

# Save the split seurat object
saveRDS(split_seurat, "C:/Users/13145/OneDrive/Desktop/BCB 5250/single_cell_rnaseq/single_cell_rnaseq/data/split_seurat.rds")

# To load the split seurat object into the environment use:
# split_seurat <- readRDS("data/split_seurat.rds")
```
